Scalable Clustering of Genotype Information using MapReduce

### 2014-05-15 doc/publications/incob2014/paper.tex
small changes to the method section

\def\includegraphic{}
\def\includegraphics{}

@@

@@

\section*{Methods}
\subsection*{Datasets}
For k-means clustering of individuals, we used variant call format (VCF) files from the 1000 Genomes Project.
These VCF files contain information about the genetic variants between each of the 1092 individuals and a reference genome
derived from GRCh37. Each of the VCF files are partitioned by chromosome.
We used the VCF file containing chromosome 1 variants (3,007,196 variants) as a subset of the entire human variant dataset.
For the entire dataset, we used each of the chromosomal VCF files. Metadata available from the 1000 Genomes Consortium
specifies additional attributes for each individual, including population and familial data.
Table~\ref{datasets} contains an overview of the datasets.

> determine in the clustering can identify ancestry 
Sequence data from hapmap samples -> determine ethnic groups
Overview see Table~\ref{datasets}

@@

\subsection*{Clustering quality}
We scored the clusters with the Adjusted Rand index. This index compares two different clusterings and assigns a value based
on their similarity. The value is zero for independent clusterings and one for identical clusterings. We compared the mahout k-means
clusters and the R k-means clusters to the known clusters based on meta data from the 1000 genome project. For the known clusters,
individuals were partitioned by their super population code (ASN, EUR, AFR, AMR). We also compared the mahout clusters to the
R clusters for the Chr1 subset data.

@@

\section*{Figures}
  \begin{figure}[h!]
  \caption{\csentence{Illustration of Mahout-based clustering of genotypes.}
      The image shows how the here introduced interface converts the vcf file to a mahout usable vector-based data type. Based on k-means it illustrates how the Map-step procedure groups of vectors by their nearest cluster center and the Reduce-step averages the vectors in each group to find the new value of the updated cluster center. The mahout-produced output is then converted into a visualisation.}
      \end{figure}

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Datasets used in this study.}
      \begin{tabular}{lcc}
        \hline
           & individuals  & variants  \\ \hline
        Chr1 subset & 1024 & 3,007,196 \\
        1000genomes trios & 1024 & ...\\
        hapmap & ... & ...  \\ \hline
      \end{tabular}
      \label{datasets}
\end{table}

\begin{table}[h!]
\caption{Adjusted Rand index (accuracy).}
      \begin{tabular}{lccc}
        \hline
           & Chr1 subset & 1000genomes trios  & hapmap \\ \hline
        Mahout, k=4 & 0.664 & ..&..\\
        Mahout, k=10 & 0.714 & ..&..\\
        Mahout, k=20 & 0.769 & .. & ..\\ 
        Mahout, k=30 & 0.728 & .. & ..\\ 
        R, k=4 & .. & - & -\\
        R, k=10 & .. & - & -\\
        R, k=20 & .. & - & -\\ 
        R, k=30 & .. & - & -\\ \hline
      \end{tabular}
      \label{datasets}
\end{table}

\begin{table}[h!]
\caption{Time taken to cluster samples (k=20).}
      \begin{tabular}{lccc}
        \hline
           & Chr1 subset & 1000genomes trios  & hapmap \\ \hline
        Mahout, 1 node & .. & ..&..\\
        Mahout, 2 nodes & .. & ..&..\\
        Mahout, 3 nodes & .. & .. & ..\\ 
        Mahout, 4 nodes & .. & .. & ..\\ 
        R & .. & - & -\\ \hline
      \end{tabular}
      \label{datasets}
\end{table}

>>>>

%\def\includegraphic{}
%\def\includegraphics{}
\usepackage{graphicx}

@@

% Our variable names
\newcommand{\chrOnePop}{Chr1\_Population}
\newcommand{\FullPop}{Full\_Population}
\newcommand{\FullFam}{Chr1\_Family}

@@

\section*{Methods}
\subsection*{Datasets}
We used variant call format (VCF) files from the 1000 Genomes Project.
These VCF files contain information about the genetic variants between each of the 1092 individuals and a reference genome derived from GRCh37. 
Each of the VCF files are partitioned by chromosome.
Metadata available from the 1000 Genomes Consortium specifies additional attributes for each individual, including population and familial data.
To be able to compare with R, we created a subset of the entire human variant dataset by using only variants from chromosome 1 (\chrOnePop) where we investigate the correct grouping of the population structure. 
The population structure is also investigated using the full dataset, i.e. all chromosomes, using Mahout libraries (\FullPop). 
In addition, we also investigate whether the full dataset allows the accurate prediction of the more detailed family association (\FullFam).
Table~\ref{datasets} contains an overview of the datasets.

@@

\subsection*{Clustering quality}
We scored the clusters with the Adjusted Rand index. 
%TODO: can you provide the formular
This index compares two different clusterings and assigns a value based on their similarity. The value is zero for independent clusterings and one for identical clusterings. We compared the Mahout k-means
clusters and the R k-means clusters to the known clusters based on meta data from the 1000 genome project. For the known clusters,
individuals were partitioned by their super population code (ASN, EUR, AFR, AMR). 
%TODO: "also on \chrOnePop" on what dataset did you compare the R performance otherwise ?
We also compared the Mahout clusters to the
R clusters for the Chr1 subset data.

@@

\section*{Figures}
  \begin{figure}[h!]
  \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{images/signature}
  \caption{\csentence{Illustration of Mahout-based clustering of genotypes.}
      The image shows how the here introduced interface converts the vcf file to a mahout usable vector-based data type. Based on k-means it illustrates how the Map-step procedure groups of vectors by their nearest cluster center and the Reduce-step averages the vectors in each group to find the new value of the updated cluster center. The mahout-produced output is then converted into a visualisation.}
      \end{figure}

@@

%% Use of \listoftables is discouraged.
%%
\section*{Tables}
\begin{table}[h!]
\caption{Datasets used in this study.}
      \begin{tabular}{lcc}
        \hline
           & individuals  & variants  \\ \hline
        \chrOnePop & 1024 & 3,007,196 \\
        \FullPop\ &  & \\
        \FullFam\ & 1024 & ...\\
      \end{tabular}
      \label{datasets}
\end{table}

\begin{table}[h!]
\caption{Adjusted Rand index (accuracy).}
      \begin{tabular}{lccc}
        \hline
           & Chr1 subset & 1000genomes trios  & hapmap \\ \hline
        Mahout, k=4 & 0.664 & ..&..\\
        Mahout, k=10 & 0.714 & ..&..\\
        Mahout, k=20 & 0.769 & .. & ..\\ 
        Mahout, k=30 & 0.728 & .. & ..\\ 
        R, k=4 & .. & - & -\\
        R, k=10 & .. & - & -\\
        R, k=20 & .. & - & -\\ 
        R, k=30 & .. & - & -\\ \hline
      \end{tabular}
      \label{datasetsAcc}
\end{table}

\begin{table}[h!]
\caption{Time taken to cluster samples (k=20).}
      \begin{tabular}{lccc}
        \hline
           & Chr1 subset & 1000genomes trios  & hapmap \\ \hline
        Mahout, 1 node & .. & ..&..\\
        Mahout, 2 nodes & .. & ..&..\\
        Mahout, 3 nodes & .. & .. & ..\\ 
        Mahout, 4 nodes & .. & .. & ..\\ 
        R & .. & - & -\\ \hline
      \end{tabular}
      \label{datasetsTime}
\end{table}
